---
title: "Ordinal regression models for Likert/rating scales data"
author: "Dr Stefano Coretta"
institute: "University of Edinburgh"
date: "2025/01/23"
format: 
  mono-light-revealjs:
    theme: [default, "custom.scss"]
    history: false
filters:
  - tachyonsextra
execute: 
  echo: true
---

```{r setup, include=FALSE}
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))

library(tidyverse)
theme_set(theme_minimal())
library(HH)
library(brms)
```

## Ordinal data

```{r}
#| label: acceptability
#| message: false

acceptability <- read_csv("https://osf.io/download/xmkvg/")
```

```{r}
#| label: acceptability-lik
acceptability_lik <- acceptability |>
  # we count the numbers for each Condition and Response combo
  count(Condition, Response) |>
  # then we pivot so that the data is wider, with columns for condition and 
  # each of the 7 values for the Response
  pivot_wider(names_from = "Response", values_from = n)
acceptability_lik
```

## Diverging stacked bar chart

```{r}
#| label: accept-likert
#| echo: false

HH::likert(
  Condition ~ .,
  acceptability_lik,
  as.percent = TRUE,
  main = "Acceptability ratings by sentence condition"
)
```

## Modelling acceptability

```{r}
#| label: acc-bm-1

acc_bm_1 <- brm(
  Response ~ Condition + (Condition | Participant) + (1 | Item),
  data = acceptability,
  family = cumulative,
  cores = 4,
  seed = 7823,
  file = "data/cache/acc_bm_1"
)
```

## Modelling acceptability: summary

```{r}
#| label: acc-bm-1-sum
summary(acc_bm_1)
```

## Modelling acceptability: regression coefficients

```{r}
#| label: acc-bm-1-fixef
fixef(acc_bm_1)
```

. . .

::: box-note
-   Each "intercept" is the estimated threshold between the Nth level and the following level, on the latent standard Gaussian distribution.

<!-- -->

-   `Conditionnoncaconical` indicates how the acceptability scores differ in the non-canonical condition relative to the canonical one on the latent Gaussian distribution. Negative values indicate lower acceptability.
:::

## Modelling acceptability: predicted acceptability

```{r}
#| label: acc-bm-1-cond

conditional_effects(acc_bm_1, categorical = TRUE)

```
